```{r setup, echo=FALSE}
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(
    echo = FALSE,
    message = FALSE,
    warning = FALSE,
    cache = TRUE
)
# options(device  = "X11")
# X11(type = "cairo")
```

```{r libraries}
library(tidyverse)
library(sf)
library(wesanderson)
library(ggplot2)
```

```{r fetch-data}
source("scripts/fetch-data.r")
```
# Introduction


## Summaries of Data sets
- overall reason for summaries and introduction to this section
- 
### Busisnesses
- background on businesses
- just relating what is noticed
- reason for data set
- describe distribution if we can
```{r businesses}
library(skimr)
library(modelsummary)

#Numberical Summary
look_at_businesses <-
  datasummary_skim(clean_businesses)
look_at_businesses

#Numberical Correlation (No categorical avaliable)
businesses_correlation <-
  datasummary_correlation(clean_businesses)
businesses_correlation

```

### Dispensaries
- background on dispensary data:
- relaying the info on what is noticed
- reason for data set
- describe distribution if we can
```{r dispensaries}
# Numerical summary
look_at_dispensaries <-
  datasummary_skim(clean_dispensaries)
look_at_dispensaries

# Numerical Correlation
dispensaries_correlation <-
  datasummary_correlation(clean_dispensaries)
dispensaries_correlation

# Categorical summaries
disp_cat <-
  datasummary_skim(clean_dispensaries, "categorical")
disp_cat
```

### Neighbourhoods
- background on neighbourhood data
- relaying the info on what is noticed
- reason for data set
- describe distribution if we can
```{r neighbourhoods}
# Categorical Data only avaliable 
look_at_neighbourhoods <-
  datasummary_skim(clean_neighbourhoods, "categorical")
look_at_neighbourhoods
```

### Intersections
- background on intersections data
- relaying the info on what is noticed
- reason for data set
- describe distribution if we can
```{r intersections}

# Numerical Summary
new_table <-
  datasummary_skim(clean_intersections)
new_table


# Categorical Summary
look_at_intersections <-
  datasummary_skim(clean_intersections, "categorical")
look_at_intersections

```

### Postal Codes
- background on postal code data
- relaying the info on what is noticed
- reason for data set
- describe distribution if we can
```{r postalcodes}
# Numerical Summary
look_at_postal_codes <-
  datasummary_skim(clean_postal_codes)
look_at_postal_codes

# Correlation table postal code
the_codes<-
  datasummary_correlation(clean_postal_codes)
the_codes

```

## Methodology
```{r extrascrubbing}
# this section is for fixing the datasets to our needs
businesses <- clean_businesses |>
    filter(`licence_address_line_3` |> 
             str_starts("M")) |>
  filter(category |> str_detect(paste(
        c(
            "RETAIL STORE \\(FOOD\\)",
            "SMOKE SHOP",
            "VAPOUR PRODUCT RETAILER"
        ),
        collapse = "|"
    )))

dispensaries <- clean_dispensaries |>
    filter(str_starts(postal_code, "M")) |>
    filter(application_status == "Authorized to Open")

```


## Visualizations
```{r}
# Cloropleth of open dispensaries by neighbourhood

dispensaries.sf <- st_as_sf(
    dispensaries,
    coords = c("Longitude", "Latitude"), crs = 4326
)

dispensaries_per_neighbourhood <- data.neighbourhoods |>
    st_contains(dispensaries.sf) |>
    lengths()

# colour code from: https://github.com/karthik/wesanderson

pal <- wes_palette("IsleofDogs2", 100, type = "continuous")

data.neighbourhoods |>
    bind_cols("dispensary_count" = dispensaries_per_neighbourhood) |>
    ggplot(aes(fill = dispensary_count)) +
    geom_sf() +
    geom_point(
        data = dispensaries,
        mapping = aes(x = Longitude, y = Latitude),
        color = "skyblue",
        size = .5,
        alpha = .5,
        inherit.aes = F
    ) +
    scale_fill_gradientn("Dispensaries", colours = pal)
   
```

## Analysis and Result




## Discussion




## Conclusion
